Realizability-Preserving Two-Moment DG Method for Fermionic Transport \tnoteref{ldrd-support}

### 2018-01-26 Documents/M1/realizableFermionicM1/sections/dg.tex
modifications to Lemma 1.  Minor edits to DG discretization.

Here we briefly outline the DG method for the moment equations.  
We do not include any physics that couple the energy dimension.  
Thus, the particle energy $\epsilonNu$ is simply treated as a parameter.  
For notational convenience, we will suppress explicit energy dependence of the moments.  
Employing Cartesian coordinates, we write the moment equations in $d$ spatial dimensions as
\begin{equation}
  \pd{\vect{\cM}}{t}+\sum_{i=1}^{d}\pderiv{}{x^{i}}\big(\,\vect{\cF}^{i}(\vect{\cM})\,\big)=\vect{\cC}(\vect{\cM}).  
  \label{eq:angularMomentsCartesian}
\end{equation}
We divide the spatial domain $D$ into a disjoint union $\cT$ of open elements $\bK$, so that $D = \cup_{\bK \in \cT}\bK$.  
We require that each element is a $d$-dimensional box in the logical coordinates; i.e.,
\begin{equation}
  \bK=\{\,\vect{x} : x^{i} \in K^{i} := (\xL^{i},\xH^{i}),~|~i=1,\ldots,d\,\}, 
\end{equation}
with surface elements denoted $\tilde{\bK}^{i}=\otimes_{j\ne i}K^{j}$.  
We let $|\bK|$ denote the volume of an element
\begin{equation}
  |\bK| = \int_{\bK}d\vect{x}, \quad\text{where}\quad d\vect{x} = \prod_{i=1}^{d}dx^{i}.  
\end{equation}
We also define the coordinates orthogonal to the $i$th dimension, $\tilde{\vect{x}}^{i}$, so that $\vect{x}=\{\tilde{\vect{x}}^{i},x^{i}\}$.  
The with of an element in the $i$th dimension is $|K^{i}|=\xH^{i}-\xL^{i}$.

We let the approximation space for the DG method, $\mathbb{V}^{k}$, to be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\bcM_{h}\in\mathbb{V}^{k}$ (which approximates $\bcM$ in Eq.~\eqref{eq:angularMomentsCartesian}) such that
\begin{align}
  &\pd{}{t}\int_{\bK}\bcM_{h}\,v\,d\vect{x}
  +\sum_{i=1}^{d}\int_{\tilde{\bK}^{i}}
  \big(\,
    \widehat{\bcF}^{i}(\bcM_{h})\,v\big|_{\xH^{i}}
    -\widehat{\bcF}^{i}(\bcM_{h})\,v\big|_{\xL^{i}}
  \,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\bcF^{i}(\bcM_{h})\,\pderiv{v}{x^{i}}\,d\vect{x}
  =\int_{\bK}\bcC(\bcM_{h})\,v\,d\vect{x},
  \label{eq:semidiscreteDG}
\end{align}
for all $v\in\mathbb{V}^{k}$ and all $\bK\in\cT$.

In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{h})$ is a numerical flux, approximating the flux on the $i$th surface of $\bK$, and is evaluated with a flux function $\widehat{\mathfrak{F}}^{i}$ using the DG approximation from both sides of the element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{h})\big|_{x^{i}}=\vect{\mathfrak{F}}^{i}(\bcM_{h}(x^{i,-},\tilde{\bx}^{i}),\bcM_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$ in the arguments of $\bcM_{h}(x^{i,-/+},\tilde{\bx}^{i})$ indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
In this paper we use the simple Lax-Friedrichs (LF) flux given by
\begin{equation}
  \vect{\mathfrak{F}}_{\mbox{\tiny LF}}^{i}(\bcA,\bcB)
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcA)+\bcF^{i}(\bcB)-\alpha^{i}\,(\,\bcA-\bcB\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux Jacobian.  
For particles propagating at the speed of light, we can simply take $\alpha^{i}=1$ (i.e., the global LF flux).

>>>>

Here we briefly outline the DG method for the moment equations.  
We do not include any physics that couple the energy dimension.  
Thus, the particle energy $\epsilonNu$ is simply treated as a parameter.  
For notational convenience, we will suppress explicit energy dependence of the moments.  
Employing Cartesian coordinates, we write the moment equations in $d$ spatial dimensions as
\begin{equation}
  \pd{\vect{\cM}}{t}+\sum_{i=1}^{d}\pderiv{}{x^{i}}\big(\,\vect{\cF}^{i}(\vect{\cM})\,\big)=\vect{\cC}(\vect{\cM}).  
  \label{eq:angularMomentsCartesian}
\end{equation}
We divide the spatial domain $D$ into a disjoint union $\mathscr{T}$ of open elements $\bK$, so that $D = \cup_{\bK \in \mathscr{T}}\bK$.  
We require that each element is a $d$-dimensional box in the logical coordinates; i.e.,
\begin{equation}
  \bK=\{\,\vect{x} : x^{i} \in K^{i} := (\xL^{i},\xH^{i}),~|~i=1,\ldots,d\,\}, 
\end{equation}
with surface elements denoted $\tilde{\bK}^{i}=\otimes_{j\ne i}K^{j}$.  
We let $|\bK|$ denote the volume of an element
\begin{equation}
  |\bK| = \int_{\bK}d\vect{x}, \quad\text{where}\quad d\vect{x} = \prod_{i=1}^{d}dx^{i}.  
\end{equation}
We also define the coordinates orthogonal to the $i$th dimension, $\tilde{\vect{x}}^{i}$, so that $\vect{x}=\{\tilde{\vect{x}}^{i},x^{i}\}$.  
The with of an element in the $i$th dimension is $|K^{i}|=\xH^{i}-\xL^{i}$.

We let the approximation space for the DG method, $\mathbb{V}^{k}$, be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\bcM_{h}\in\mathbb{V}^{k}$ (which approximates $\bcM$ in Eq.~\eqref{eq:angularMomentsCartesian}) such that
\begin{align}
  &\pd{}{t}\int_{\bK}\bcM_{h}\,v\,d\vect{x}
  +\sum_{i=1}^{d}\int_{\tilde{\bK}^{i}}
  \big(\,
    \widehat{\bcF}^{i}(\bcM_{h})\,v\big|_{\xH^{i}}
    -\widehat{\bcF}^{i}(\bcM_{h})\,v\big|_{\xL^{i}}
  \,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\bcF^{i}(\bcM_{h})\,\pderiv{v}{x^{i}}\,d\vect{x}
  =\int_{\bK}\bcC(\bcM_{h})\,v\,d\vect{x},
  \label{eq:semidiscreteDG}
\end{align}
for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathscr{T}$.

In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{h})$ is a numerical flux, approximating the flux on the surface of $\bK$ with unit normal along the $i$th coordinate direction, and is evaluated with a flux function $\vect{\mathscr{F}}^{i}$ using the DG approximation from both sides of the element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{h})\big|_{x^{i}}=\vect{\mathscr{F}}^{i}(\bcM_{h}(x^{i,-},\tilde{\bx}^{i}),\bcM_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$ in the arguments of $\bcM_{h}(x^{i,-/+},\tilde{\bx}^{i})$ indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
In this paper we use the simple Lax-Friedrichs (LF) flux given by
\begin{equation}
  \vect{\mathscr{F}}_{\mbox{\tiny LF}}^{i}(\bcA,\bcB)
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcA)+\bcF^{i}(\bcB)-\alpha^{i}\,(\,\bcA-\bcB\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux Jacobian.  
For particles propagating at the speed of light, we can simply take $\alpha^{i}=1$ (i.e., the global LF flux).

### 2018-01-26 Documents/M1/realizableFermionicM1/sections/realizability.tex
modifications to Lemma 1.  Minor edits to DG discretization.

@@

\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they are consistent with a distribution function satisfying $0\le f \le 1$.
\end{define}
\begin{lemma}
  Let the distribution function $f$ be bounded by $0\le f \le1$.  
  Then $0\le\cJ\le1$ and $\big(1-\cJ\big)\cJ-|\vect{\cH}|\ge0$
\end{lemma}
\begin{proof}
  Since this lemma is independent from $\vect{z}$ and t, we can drop them from the expression for convenience.
  Let's proof $0\le\cJ\le1$.

  Since $f$ is positive, we have
    \begin{equation}
      \cJ=\f{1}{4\pi}\int_{\bbS^{2}}f(\omegaNu)\,d\omegaNu\ge0.
    \end{equation}

  Since $1-f$ is positive, we have
    \begin{equation}
      1-\cJ=\f{1}{4\pi}\int_{\bbS^{2}}(1-f(\omegaNu))\,d\omegaNu\ge0.
    \end{equation}
    Therefore, $0\le\cJ\le1$.

    For $\big(1-\cJ\big)\cJ-|\vect{\cH}|\ge0$, we can proof it's valid with 1-dimension $\vect{\cH}$ first.
    In 1-dimension, $\vect{\ell}(\omegaNu) = \omegaNu$ and
    \begin{equation}
    \vect{\cH}
      =\f{1}{4\pi}\int_{\bbS^{2}}f(\omegaNu) \omegaNu d\omegaNu = \f{1}{2}\int_{-1}^{1}f(\omegaNu) \omegaNu d\omegaNu.
    \end{equation}
    Following Banach \& Larecki \cite{banachLarecki_2017}, we can rewrite it to be
    \begin{align}
    \vect{\cH}
        &=(1-\cJ)\,\cJ \nonumber \\
        &-\Big\{\,
          \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\omegaNu)f(\omegaNu)\,d\omegaNu
          +\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\omegaNu)[1-f(\omegaNu)]\,d\omegaNu
        \,\Big\}.
        \label{eq:momentumRewrite1}
    \end{align}
    Since $(1-2\cJ-\omegaNu)f(\omegaNu)$ is positive when $-1\leq\omegaNu\leq1-2\cJ$ and $(2\cJ-1+\omegaNu)[1-f(\omegaNu)]$ is positive when $1-2\cJ\leq\omegaNu\leq1$,
    \begin{align}
    \vect{\cH} \leq (1-\cJ)\,\cJ.
    \label{eq:HupBry1D}
    \end{align}
    Similarly, we can also rewrite $\vect{\cH}$ as
    \begin{align}
     \vect{\cH}
        &=-(1-\cJ)\,\cJ \nonumber \\
        &-\Big\{\,
          \f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\omegaNu)[1-f(\omegaNu)]\,d\omegaNu
          +\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\omegaNu)f(\omegaNu)\,d\omegaNu
        \,\Big\}.
        \label{eq:momentumRewrite2}
    \end{align}
    Since $(1-2\cJ+\omegaNu)[1-f(\omegaNu)]$ is negative when $-1\leq\omegaNu\leq2\cJ-1$ and $(2\cJ-1-\omegaNu)f(\omegaNu)$ is negative when $2\cJ-1\leq\omegaNu\leq1$,
    \begin{align}
    \vect{\cH} \geq -(1-\cJ)\,\cJ.
     \label{eq:HlowBry1D}
    \end{align}
    Combining Eq.~\eqref{eq:HupBry1D} and ~\eqref{eq:HlowBry1D} gives $\big(1-\cJ\big)\cJ-|\vect{\cH}|\ge0$ in 1-dimension.
\end{proof}

We now define the realizable set $\cR$
\begin{equation}
  \cR:=\big\{\,\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}~|~\cJ\in[0,1]~\text{and}~(1-\cJ)\cJ-|\vect{\cH}|\ge0\,\big\}
  \label{eq:realizableSet}
\end{equation}

\begin{lemma}
  The set $\cR$, defined in Eq.~\eqref{eq:realizableSet}, is convex.  
\end{lemma}
\begin{proof}
  Prove this
\end{proof}

>>>>

Since we only consider the angular dependence of the distribution function in this section, we simplify the notation by suppressing the $\vect{z}$ and $t$ dependence, and write $f(\omegaNu,\vect{z},t)=f(\omegaNu)$.

@@

\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they are consistent with a distribution function satisfying $0\le f(\omegaNu) \le 1$.
\end{define}

We begin by stating the bounds on the moments by restating results from Theorem~7.1 in \cite{banachLarecki_2017} in the following Lemma.  
\begin{lemma}
  Let the distribution function be bounded by $0\le f(\omegaNu) \le1$ for all $\omegaNu\in\bbS^{2}$.  
  Then $0\le\cJ\le1$ and $\big(1-\cJ\big)\,\cJ-|\vect{\cH}|\ge0$.  
\end{lemma}
\begin{proof}
  First, let $\mu=\cos\thetaNu$ and write
  \begin{equation}
    \cJ=\f{1}{2}\int_{\bbS^{1}}\mathfrak{f}(\mu)\,d\mu,
    \quad\text{where}\quad
    \mathfrak{f}(\mu)=\f{1}{2\pi}\int_{0}^{2\pi}f(\mu,\phiNu)\,d\phiNu.  
  \end{equation}  
  Since $f(\omegaNu)\in[0,1]$, we have $\mathfrak{f}(\mu)\in[0,1]$, which implies $\cJ\in[0,1]$.  
  Next we need to show that $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
  Let $\widehat{\vect{n}}\in\bbR^{3}$ (independent of $\omegaNu$) be an arbitrary unit vector, and write
  \begin{equation}
    \widehat{\vect{n}}\cdot\vect{\cH}
    =\int_{\bbS^{1}}\mathfrak{f}(\mu)\,\mu\,d\mu,
    \label{eq:NdotH}
  \end{equation}
  where we have aligned the polar axis of the spherical momentum space coordinate system with $\widehat{\vect{n}}$; i.e., $\widehat{\vect{n}}\cdot\vect{\ell}(\omega)=\mu$.  
  First, following \cite{banachLarecki_2017}, we can write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    (1-\cJ)\cJ - \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\mu)\,\mathfrak{f}(\mu)\,d\mu
    -\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu.  
  \end{equation*}
  Since $\mathfrak{f}\in[0,1]$, $(1-2\cJ-\mu)$ is nonnegative on the interval $\mu\in[-1,1-2\cJ]$, and $(2\cJ-1+\mu)$ is nonnegative on the interval $\mu\in[1-2\cJ,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  In a similar manner, we can also write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    -(1-\cJ)\cJ-\f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu
    -\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\mu)\,\mathfrak{f}(\mu)\,d\mu.  
  \end{equation*}
  Then, since $\mathfrak{f}\in[0,1]$, $(1-2\cJ+\mu)$ is nonpositive on the interval $\mu\in[-1,2\cJ-1]$, and $(2\cJ-1-\mu)$ is nonpositive on the interval $\mu\in[2\cJ-1,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\ge-(1-\cJ)\,\cJ$.  
  We therefore have $-(1-\cJ)\,\cJ\le\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  Since this holds for any unit vector $\widehat{\vect{n}}\in\bbR^{3}$, we must have $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
\end{proof}

We now define the realizable set $\cR$
\begin{equation}
  \cR:=\big\{\,\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}~|~\cJ\in[0,1]~\text{and}~\gamma(\vect{\cM})\equiv(1-\cJ)\cJ-|\vect{\cH}|\ge0\,\big\}
  \label{eq:realizableSet}
\end{equation}

\begin{lemma}
  The set $\cR$, defined in Eq.~\eqref{eq:realizableSet}, is convex.  
\end{lemma}
\begin{proof}
  A convex set $Q$ is a set has the following property: 
  \begin{align}
  \theta a + (1-\theta)b \in Q, \text{\space} \text{for } \forall a, b \in Q \text{ and } 0\leq\,\theta\,\leq1. 
  \end{align} 
  Let $\vect{\cM}_{a}=\big(\cJ_{a},\vect{\cH}_{a}\big)^{T}$ and $\vect{\cM}_{b}=\big(\cJ_{b},\vect{\cH}_{b}\big)^{T}$ to be two arbitrary elements of $\cR$.
  If $\cR$ is convex, we will have $\vect{\cM}_{ab} = \big(\cJ_{ab},\vect{\cH}_{ab}\big)^{T} = \theta \vect{\cM}_{a} + (1-\theta)\vect{\cM}_{b}$ in $\cR$ with $0\leq\theta\leq1$.
  Let's prove it is true.

  For the first component
  \begin{align}
  \cJ_{ab} = \theta \cJ_{a} + (1-\theta)\cJ_{b},
  \end{align}
  since $\theta$, $\cJ_{a}$, $1-\theta$ and $\cJ_{b}$ are all nonnegative, $\cJ_{ab}$ is also nonnegative ($\cJ_{ab}\geq0$).
  Meanwhile, since $\cJ_{a}\leq1$ and $\cJ_{b}\leq1$, 
  \begin{align}
  \cJ_{ab} \leq \theta \cdot 1 + (1-\theta) \cdot 1 = 1.
  \end{align}
  Therefore, $\cJ_{ab} \in [0,1]$.

  For the second component,
  \begin{align}
  \vect{\cH}_{ab} 
  & = \theta \vect{\cH}_{a} + (1-\theta)\vect{\cH}_{b} \nonumber\\
  & = \f{1}{4\pi}\int_{\bbS^{2}} \left[ \theta\,f_{a}(\omegaNu) + (1-\theta) f_{b}(\omegaNu) \right] \bl(\omegaNu) d\omegaNu. 
  \end{align}
  Since $\theta\,f_{a}(\omegaNu) + (1-\theta) f_{b}(\omegaNu) \in [0,1]$, we can define $\tilde{f} = \theta\,f_{a}(\omegaNu) + (1-\theta) f_{b}(\omegaNu)$.
  $\tilde{f} \ in [0,1]$ and 
  \begin{align}
  \cJ_{ab} = \f{1}{4\pi}\int_{\bbS^{2}} \tilde{f} d\omegaNu \\
  \vect{\cH}_{ab} = 
  \end{align}
\end{proof}
